---
title: "Lecture 12: Lists, Iterations, and `purrr`"
subtitle: "November 25, 2025"
date-modified: last-modified
---

### **Learning Objectives**

From this topic, students are anticipated to be able to:

-   Use the `map` family of functions from the purrr package to iteratively apply a function.

-   Create and operate on list columns in a tibble using `nest()`, `unnest()`, and the `map_*()` family of functions.

-   Define functions on-the-fly within a `map` function using shortcuts.

-   Create and operate on list columns in a tibble using `nest()`, `unnest()`

-   Apply list columns to cases in data analysis: columns of models, columns of nested lists (JSON-style data), and operating on entire groups within a tibble.

We will need to load in the `tidyverse`  package for this lecture:

```{r, warning = F, message = F}
library(tidyverse)
```


## Lecture Slides

-   [Lecture 13 - Lists, Iteration, and `purrr`](../slides/slidefiles/lec13_purrr_slides.qmd)


## Lists

Here is a list in R; it holds multiple items. 

```{r}
my_list <- list(1:10, c("a", "b", "c"), "samplestring")
my_list
```

A list might sound like a *vector*, which we have worked with before – remember, we construct them using the `c()` function. Indeed, vectors and lists can both hold multiple items. But there are key differences.

| Vectors | Lists |
|----|----|
| Access elements with square brackets `[]` | Access elements with `[[]]` |
| Each element must be an atomic data type (i.e., a single value) | Elements can be anything, even another list or another vector |
| Each element has to be of the same type | Elements can be as different as you like |


Let's take our sample list and access some items stored in it. 

```{r}
#returns the vector stored in the second element of the list
my_list[[2]] 
```

To access the data within the vector, we can index it as well:

```{r}
#returns third element of the vector stored in the second element of the list
my_list[[2]][3] 
```

Elements within a list can also be named using `names()`.

```{r}
names(my_list) <- c("NumericVec", "CharacterVec", "String")
my_list
```

Once the elements are named, you can access them using the `$` operator, similar to how you can grab columns from a data frame or tibble:

```{r}
my_list$CharacterVec
```




## Tibbles, as a list?

Speaking of data frames and tibbles, did you know that data frames and tibbles are actually a special type of list? It’s true!

```{r}
typeof(mtcars) 
```

```{r}
typeof(palmerpenguins::penguins)
```

It turns out that they are actually *lists*, where each element of the list stores a column, which is either a list with the same number of entries as the tibble has rows, or a vector with the same number of entries as the tibble has rows.

This has an important implication: we can efficiently apply a function to each column of a tibble by learning how to apply a function to each entry of a list. This is yet another way (beyond functions themselves) of avoiding duplicating code, which you will recall (from the functions topic) has many advantages.

## Iteration

If you programmed before, you probably have an idea of how to do this with a `for` loop. Here’s an example of a for loop in R that iterates over the entries of a numeric vector `x`, squares each entry, and stores the result in a numeric vector `output`:

```{r}
x <- 1:10 
output <- vector("double", length(x)) #init a vector of 0s

for(i in seq_along(x)) { 
    output[i] <- x[i]^2  
}

output
```


## Iteration with `purrr`

Often, you can replace loops with a compact call to a function in the `purrr` package. This has the advantage of making our code even more readable and compact, since we’re expressing the same logic with less space. 

There are many `map_*()` functions, a few of which we will highlight here:

- `map()`: applies a function to each element in a list or vector, returns a list
- `map_dbl()`: applies a function to each element in a list or vector, returns a vector
- `map_dfr()`: applies a function to each element in a list or vector, returns a data frame
- `map2()`: performs `map()` over multiple inputs simultaneously


The first argument of each function specifies the list/vector we want to iterate over, and the second argument specifies a function that we want to apply to each entry.

Let's use these map functions over a list containing ages of some made-up people

```{r}
sample_list <- list(18, 25, 63, 22)
names(sample_list) <- c("Amir", "Jenna", "Logan", "Phum")

sample_list
```

We will compare the outputs of these functions when applying a simple square function to their ages. 

### `map()`

```{r}
map(sample_list, function(x) x^2) #map a square function to the sample vector
```

A list is returned!

### `map_dbl()`

```{r}
map_dbl(sample_list, function(x) x^2) #map a square function to the sample vector
```

A vector is returned!


### `map_dfr()`

```{r}
map_dfr(sample_list, function(x) x^2) #map a square function to the sample vector
```

A dataframe is returned!


You will explore `map2()` in the Worksheet. 

## `purrr` Shortcuts

Here’s an example using `purrr::map_dbl()` and a custom function:

```{r}
purrr::map_dbl(sample_list, function(x) x^2)
```

 Options for specifying functions include the name of a function, a fully specified custom function (as demonstrated above), or one of the “shortcuts” the `purrr` developers have provided.

Here are two examples of “shortcuts”:

```{r}
purrr::map_dbl(sample_list, ~ (.x)^2)
```

```{r}
purrr::map_dbl(sample_list, \(x) x^2)
```

The second one is easier to remember and appears to be the one that `purrr` developers are recommending now; see the [purrr cheatsheat](https://github.com/rstudio/cheatsheets/blob/main/purrr.pdf). But this change in recommendation appears to have happened around 2022/2023, so you may still see the first type of shortcut in many places in the wild.




## List Columns

Did you know columns in a tibble can have type “list”? We call these types of columns “list columns”.

Consider the following example: a snippet of the Game of Thrones data from [An API of Ice and Fire](https://anapioficeandfire.com/).

```         
## # A tibble: 6 × 3
##   name              gender titles   
##   <chr>             <chr>  <list>   
## 1 Theon Greyjoy     Male   <chr [2]>
## 2 Tyrion Lannister  Male   <chr [2]>
## 3 Victarion Greyjoy Male   <chr [2]>
## 4 Will              Male   <chr [1]>
## 5 Areo Hotah        Male   <chr [1]>
## 6 Chett             Male   <chr [1]>
```

Some characters have one title (e.g., Will); others have more than one title (e.g., Theon Greyjoy). Consequently, the `titles` column is a *list column*, where each entry is a *list* that contains as many or as few strings as we like.

We can even have tibbles as an observation in our data (which is technically still a list type). 

Nested columns can have lists nested in lists, which we refer to as JSON-style data. You'll see this in Worksheet B3!

### Create List Columns using `nest()`

```{r}
# Artificial dataset of families
family_data <- tibble(
  family_id = c(1, 1, 2, 3, 3, 3, 4),
  lastname    = c("Smith", "Smith", "Lee", "Patel", "Patel", "Patel", "LeBlanc"),
  child     = c("Mia", "Leo", "Noah",
                "Ella", "Sofia", "Lucas",
                "Harper")
)

family_data
```

Instead of having one row per child, perhaps we want to nest the data such that there is only one row per family. We can nest the `child` column! (*Aside: in general this data will not be tidy, but we're demonstrating how to nest and un-nest data here*).

We first need to group by `family_id` to gather children from each family, and then create a nested column called "children":

```{r}
nested_families <- family_data %>%
  group_by(family_id, lastname) %>%   # group by family
  nest(children = child) %>%           # create list-column "children" (tibble type)
  mutate(children = map(children, ~ .x$child)) #overwrite children: change from tibble to vector

nested_families
```

Now we have a list column that contains vectors of varying lengths with the children's names within each family. 


### Remove List Columns using `unnest()`

We can revert back to the original tibble using `unnest()`. 

```{r}
original_family_data <- nested_families %>%
  unnest(children) #unnest the children column name

original_family_data
```





## Resources

Video lectures:

-   [Vectors and Lists in R](https://www.youtube.com/watch?v=nJgrDoCENBI)

-   [R List Columns: purrr map and nesting](https://www.youtube.com/watch?v=_731PZ8I9m0)

Written material:

-   [R4DS Chapter 21: Iteration](https://r4ds.hadley.nz/iteration.html) for purrr

    -   21.1 for an intro

    -   21.5 for the `map` family of functions

    -   The intro of 21.7 for the `map2` and `pmap` families.

-   [“List Columns”](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html) from [Jenny’s purrr tutorial](https://jennybc.github.io/purrr-tutorial/index.html)

-   [“Nested data” article](https://tidyr.tidyverse.org/articles/nest.html) on tidyr’s website.

Want to dig deeper? These resources can help.

-   [Advanced R Chapter 9: Functionals](https://adv-r.hadley.nz/functionals.html) – looking at purrr and `map()` from a programming perspective.

-   [tidyr’s rectangling vignette](https://tidyr.tidyverse.org/articles/rectangle.html) – for handling deeply nested lists (JSON-style data), similar to tidyr’s `pivot_` functions.
